Finding Maximum $k$-Cliques Faster using Lazy Global Domination

### 2016-04-11 paper/paper.tex
Tweaking

A clique in a graph is a set of vertices, each of which is adjacent to every other vertex in the
set. Finding a clique of maximum size in a graph is one of the basic \NP-hard problems
\cite{Garey:1990}; applications include geometry, coding theory, computer vision and bioinformatics
\cite{Bomze:1999,Butenko:2006}. However, when analysing real-world data, a clique may be too strong
a requirement. A $k$-clique\footnote{Or sometimes $n$-clique or $s$-clique; in an unfortunate clash
of notation, ``$k$-clique problem'' is sometimes used elsewhere for the decision version of the
clique problem, to distinguish it from the maximum clique problem.} is a relaxed form of clique,
where instead of requiring each pair of vertices to be directly adjacent, we only require they be
connected by a path of length at most $k$ \cite{Luce:1950}.  Thus a $1$-clique \emph{is} a clique, a
$2$-clique may be thought of as ``a group of people, all of whom either know each other or have a
mutual acquaintance'', and so on. We illustrate this in \cref{figure:k-cliques}.  Determining the
size of a maximum $k$-clique is \NP-hard for any fixed $k$ \cite{Bourjolly:2002}.

@@

Let $v$ and $w$ be distinct vertices in a graph $G$ (they may or may not be adjacent). We say that
$v$ \emph{dominates} $w$ if the neighbourhood of $w$, excluding $v$, is a (possibly non-strict)
subset of the neighbourhood of $v$, excluding $w$. From a maximum clique perspective, this means
that $v$ is ``better than'' $w$. If $v$ and $w$ are adjacent, any clique containing $w$ may always
be extended by the inclusion of $v$; if $v$ and $w$ are non-adjacent, replacing $w$ with $v$ in any
clique containing $w$ cannot reduce the amount by which the clique may be grown.

@@

Seven of these graphs were considered for the $k$-club problem by Wotzlaw \shortcite{Wotzlaw:2014}. In
these cases, the $2$-clique and $2$-club numbers are the same, except for ``football'' where the
$2$-club number is 16 but the $2$-clique number is 17; for $k = 3$ and $k = 4$ there are some
differences. The difference in computational difficulty between the $k$-clique and $k$-club problems
really stands out here: for ``polblogs'' with $k = 3$ and $k = 4$, Wotzlaw was unable to prove
optimality within an hour, but we required less than two seconds to do so. In both of these cases the
$k$-clique and $k$-club numbers are the same.

@@

Parallel search is an active research area in maximum clique algorithms
\cite{DBLP:journals/algorithms/McCreeshP13,DBLP:journals/jcisd/DepolliKRTJ13,DBLP:journals/topc/McCreeshP15,DBLP:journals/cor/SegundoLP16}.
The idea is to work with a shared incumbent, and to speculatively explore portions of the search
tree in parallel using multiple threads, in the hopes that most of the speculative work will
contribute to the optimality proof.

We selected the four graphs which we were unable to solve (for some values of $k$), along with two
of the challenging graphs which required substantial amounts of search (at least half a million
nodes) to solve, and repeated the experiments using parallel search with 20 threads, a timeout of 12
hours, and the domination rule (with the laziness made thread-safe) enabled. The results are shown
in \cref{table:parallel}.  Parallel search allowed us to close one more instance (``cond-mat'' with
$k=3$), and improved the bounds on ??  further instances. However, to get sufficient work balance,
we had to use the resplitting strategy proposed by \citeauthor{DBLP:journals/topc/McCreeshP15}
\shortcite{DBLP:journals/topc/McCreeshP15}, and we had
to allow for work splitting to depth 10 rather than depth 3. Simpler static
decomposition approaches \cite{DBLP:journals/jcisd/DepolliKRTJ13,DBLP:journals/cor/SegundoLP16}
would give little to no speedup on many harder $k$-clique instances, unless it were somehow possible
to generate $\mathcal{O}(|V|^{10})$ subproblems.

@@

On the other hand, for ``wing-nodal'' with $k=4$, the parallel run did much less work than the
sequential run: a speedup of over 100 was obtained from 20 cores. A similar effect occurred with
``cond-mat'' and $k=4$, and on several of the graphs where the incumbent was improved, it was
improved after less time than the sequential runs took. This is because the work splitting mechanism
we used explicitly diversifies at the top of search first, where branching heuristics are least
likely to be correct, which leads to an initial incumbent being found faster---this is in line with
recent observations that tailored work stealing should be favoured over randomised work stealing for
combinatorial search problems \cite{DBLP:conf/cp/ChuSS09,DBLP:journals/topc/McCreeshP15}.

@@

We have shown that using a maximum clique algorithm to solve the maximum $k$-clique algorithm for a
graph $G$ by considering $G^k$ in place of $G$ is feasible in practice. This is despite $G^k$
potentially being dense even if $G$ is sparse.

@@

>>>>

A clique in a graph is a set of vertices, each of which is adjacent to every other vertex in the
set. Finding a clique of maximum size in a graph is one of the basic \NP-hard problems
\cite{Garey:1990}; applications include geometry, coding theory, computer vision and bioinformatics
\cite{Bomze:1999,Butenko:2006}. However, when analysing real-world data, a clique may be too strong
a requirement. A $k$-clique (or sometimes $n$-clique or $s$-clique---in an unfortunate clash of
notation, ``$k$-clique problem'' is sometimes used elsewhere for the decision version of the clique
problem, to distinguish it from the maximum clique problem) is a relaxed form of clique, where
instead of requiring each pair of vertices to be directly adjacent, we only require they be
connected by a path of length at most $k$ \cite{Luce:1950}.  Thus a $1$-clique \emph{is} a clique, a
$2$-clique may be thought of as ``a group of people, all of whom either know each other or have a
mutual acquaintance'', and so on. We illustrate this in \cref{figure:k-cliques}.  Determining the
size of a maximum $k$-clique is \NP-hard for any fixed $k$ \cite{Bourjolly:2002}.

@@

Let $v$ and $w$ be distinct vertices in a graph $G$. We say that $v$ \emph{dominates} $w$ if the
neighbourhood of $w$, excluding $v$, is a (possibly non-strict) subset of the neighbourhood of $v$,
excluding $w$. From a maximum clique perspective, this means that $v$ is ``better than'' $w$. If $v$
and $w$ are adjacent, any clique containing $w$ may always be extended by the inclusion of $v$; if
$v$ and $w$ are non-adjacent, replacing $w$ with $v$ in any clique containing $w$ cannot reduce the
amount by which the clique may be grown.

@@

Seven of these graphs were considered for the $k$-club problem by Wotzlaw \shortcite{Wotzlaw:2014}.
In these cases, the $2$-clique and $2$-club numbers are the same, except for ``football'' where the
$2$-club number is 16 but the $2$-clique number is 17; for $k = 3$ and $k = 4$ there are some
differences. There is a large difference in computational difficulty between the $k$-clique and
$k$-club problems: for ``polblogs'' with $k = 3$ and $k = 4$, Wotzlaw was unable to prove optimality
within an hour, but we required less than two seconds to do so. In both of these cases the
$k$-clique and $k$-club numbers are the same.

@@

Parallel search is an active research area for maximum clique algorithms
\cite{DBLP:journals/algorithms/McCreeshP13,DBLP:journals/jcisd/DepolliKRTJ13,DBLP:journals/topc/McCreeshP15,DBLP:journals/cor/SegundoLP16}.
The idea is to work with a shared incumbent, and to speculatively explore portions of the search
tree in parallel using multiple threads, in the hopes that most of the speculative work will
contribute to the optimality proof.

We selected the four graphs which we were unable to solve (for some values of $k$), along with two
of the challenging graphs which required substantial amounts of search (at least half a million
nodes) to solve, and repeated the experiments using parallel search with 20 threads and the
domination rule (with the laziness made thread-safe) enabled, using the resplitting strategy
described by \citeauthor{DBLP:journals/topc/McCreeshP15} \shortcite{DBLP:journals/topc/McCreeshP15}.
The results are shown in \cref{table:parallel}. Parallel search allowed us to close two more
instances, and improved the bounds on ?? further instances with 12 hours of search. However, to get
sufficient work balance, we had to allow for work splitting to depth 10 rather than depth 3. A close
inspection of the search patterns showed that simpler static decomposition approaches
\cite{DBLP:journals/jcisd/DepolliKRTJ13,DBLP:journals/cor/SegundoLP16} would give little to no
speedup on many harder $k$-clique instances, unless it were somehow possible to generate
$\mathcal{O}(|V|^{10})$ subproblems.

@@

On the other hand, for ``wing-nodal'' with $k=4$, the parallel run did much less work than the
sequential run: a speedup of over 100 was obtained from 20 cores. A similar effect occurred with
``cond-mat'' and $k=4$, and on ?? of the graphs where the incumbent was improved, it was improved
after less time than the sequential runs took. This is because the work splitting mechanism we used
explicitly diversifies at the top of search first, where branching heuristics are least likely to be
correct, which leads to an initial incumbent being found faster---this is in line with recent
observations that tailored work stealing should be favoured over randomised work stealing for
combinatorial search problems \cite{DBLP:conf/cp/ChuSS09,DBLP:journals/topc/McCreeshP15}.

@@

We have shown that using a maximum clique algorithm to solve the maximum $k$-clique algorithm for a
graph $G$ by considering $G^k$ in place of $G$ is feasible in practice. This is despite $G^k$
potentially being dense even if $G$ is sparse---this ruled out the use of maximum clique algorithms
which are designed for sparse graphs
\cite{DBLP:conf/waw/PattabiramanPGLC13,DBLP:journals/cor/SegundoLP16}, and we were considering
graphs with many more vertices than is typical for dense maximum clique algorithms.

@@

We were able to use parallelism to close two further instances, although this required a much finer
level of task granularity than usual.  We suspect further progress could be made, for example, by
tailoring the initial vertex ordering based upon what we know about the graphs, or by increasing the
number of recursive calls per second by making the colouring stage cheaper, for example by reusing
colour classes \cite{DBLP:conf/lion/NikolaevBS15}.
